---
title: "Statistics 5014: Homework 3"
<<<<<<< HEAD
subtitle: "Due Monday September 19, 10 am"
=======
subtitle: "Due Wednesday, September 12, 10 am"
>>>>>>> cba3a8790023f1e10c2be379469397faa3443b8f
date: '`r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(dplyr)
library(knitr)
library(plotrix)
library(tidyr)
```

## Problem 4

Coding is akin to writing literature, i.e, using correct punctuation. It is important to be consistent and concise throughout the text, so that reviewing and debugging can be performed efficiently. I will attempt to implement the nomenclature scheme that the style guides suggests to be more consistent in naming and differentiating my variables and functions.

## Problem 5: lint HW2

```{r eval=F, echo=T}
    lint(filename = "./02_data_munging_summarizing_R_git/HW2_Park_Stephen.Rmd")
```

As a result of linting HW2, the following issues were brought up:

- Excessively long lines ( > 80 characters),
- inconsistent spacing around infix operators and commas,
- use of single quotes over double-quotes, and
- variable and function names that are not all lowercase.

The general message from the linting process is that the code syntax needs to be consistent and arranged in a manner so that it is easy to review. The following changes were made to the existing code:

- Excessively long commands and/or comments were divided so that each line did not occupy more than 80 characters.
- Spaces were included before and after infix operators, and after commas.
- All single quotes were substituted with double quotes.
- Only lowercase was used for variable and function names.

\newpage
## Problem 6

```{r}
dev.input <- readRDS("HW3_data.rds")   # Read in dataset from RDS file

# Define function to display descriptive statistics
DevStats <- function(input){
  obs.count <- input %>% count(Observer)  # Frequency of each Observer
  # Create empty output table with target variables
  output <- data.frame(id = obs.count[, 1],
                            mean.dev1 = numeric(nrow(obs.count)),
                            mean.dev2 = numeric(nrow(obs.count)),
                            sd.dev1 = numeric(nrow(obs.count)),
                            sd.dev2 = numeric(nrow(obs.count)),
                            corr = numeric(nrow(obs.count)))
# Group data by Observer and calculate descriptive statistics
  for (i in 1:nrow(obs.count)){
    sub.data <- filter(input, Observer == i)
    output$mean.dev1[i] <- mean(sub.data[, 2])
    output$mean.dev2[i] <- mean(sub.data[, 3])
    output$sd.dev1[i] <- sd(sub.data[, 2])
    output$sd.dev2[i] <- sd(sub.data[, 3])
    output$corr[i] <- cor(sub.data[, 2], sub.data[, 3])
  }
  return(output)
}
dev.output <- DevStats(dev.input)   # Run function with given data

# Create table for output data
kable(dev.output, digits = 3, format = "pandoc",
      caption = "Descriptive Statistic Summary of Device Measurements",
      col.names = c("Observer", "Dev1 Mean", "Dev2 Mean", "Dev1 SD",
                    "Dev2 SD", "Correlation"))
# Configure boxplot layout
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE),
   widths=c(2,2), heights=lcm(c(4,6)))
par(mar=c(4, 4, 1, 1))
# Side-by-side boxplot to compare means
boxplot(dev.output$mean.dev1, dev.output$mean.dev2, main="Means by Device",
        ylab="Measurement", ylim = c(45, 55), names = c("Dev1", "Dev2"))
# Detailed boxplots to view spreads
boxplot(dev.output$mean.dev1, xlab = "Dev1", ylab = "Measurement",
        ylim = c(54.26, 54.27))
boxplot(dev.output$mean.dev2, xlab = "Dev2", ylab="Measurement",
        ylim = c(47.83, 47.84))
# Configure violin plot layout
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE),
   widths=c(2,2), heights=lcm(c(4,6)))
par(mar=c(4, 4, 1, 1))
# Side-by-side violin plot to compare means
violin_plot(data.frame(dev.output$sd.dev1, dev.output$sd.dev2), main = "Standard Deviations by Device", ylab = "Measurement", x_axis_labels = c("Dev1", "Dev2"))
# Detailed boxplots to view spreads
violin_plot(dev.output$mean.dev1, x_axis_labels = "Dev1", ylab = "Measurement", main = NA)
violin_plot(dev.output$mean.dev2, x_axis_labels = "Dev2", ylab = "Measurement", main = NA)
```

## Problem 7

```{r, tidy.opts = list(width.cutoff=70), tidy=TRUE}
# Store URL for blood pressure data
url <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/BloodPressure.dat"
 # Import data; use 2nd row as header
data_mess <- fread(url, skip = 1, header = T, fill = T, sep = " ")
# Preserve messy data and manipulate using a new variable
data_tidy <- data_mess %>% gather(key = Reader, value = Blood.Pressure) %>%
  filter(Reader != "Day") %>% arrange(Day)
head(data_tidy, 7)   # Preview data set
summary(data_tidy)   # Summarize data
str(data_tidy)       # Display blood pressure data structure
```

## Problem 8: Newton's Method

```{r}
# Pre-define function
func <- function(x) {
  3^x - sin(x) + cos(5*x)
}
# Define Newton's Method function with the following input:
# f = given funtion
# x_0 = initial x
# tol = tolerance
# n = max number of iterations
newton <- function(f, x_0, tol = 1e-5, n = 1000) {
  require(numDeriv)
  # If x_0 is zero, end function and output x_0
  if (f(x_0) == 0.0) {
    return(x_0)
  }
  x <- x_0
  for (i in 1:n) {
    x[i+1] <- x[i] - f(x[i])/genD(func = f, x = x[i])$D[1]
    if (x[i+1] - x[i] < tol) {
      print("Convergence achieved")
      break
    }
    if(i == n) {
      print("Method did not converge")
    }
  }
  print(paste0("The tolerance for convergence is ", tol))
  plot(x[1:length(x) - 1], f(x[1:length(x) - 1]), col = "blue", ylab = "f(x)", xlab = "x", xlim = c(-max(abs(x)),max(abs(x))), ylim = c(-max(abs(f(x))),max(abs(f(x)))))
  points(x[length(x)], f(x[length(x)]), col = "red")
  abline(h=0)
  abline(v=0)
  x.data <- tibble::rowid_to_column(data.frame(x), "Iterations")
  kable(x.data, digits = 3, format = "pandoc",
      caption = "Iterative Estimations in Newton's Method")
}
# Test the method with given function
newton(func, -10, tol = 1e-6)
```
