---
title: "Statistics 5014: Homework 2"
subtitle: "Due Monday September 11, 10 am"
author: "Stephen Park"
date: '`r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = "")
library(data.table)
library(dplyr)
library(tidyr)
library(swirl)
library(knitr)
library(kableExtra)
```
  
## Problem 4: How would version control help in the classroom?

When writing long script files with complex codes, e.g., LaTeX code lines that need to be knitted to see their output, version control would make it efficient to backtrack to any minor errors that may have been made and correct them. The presence of a version control system would make myself more bold in attempting new coding techniques and syntaxes as well. It will also be more easy to collaborate on projects with fellow classmates without having to be physically in the same work area.

## Problem 5: Create tidy datasets from Wu and Hamada (2009)

a. Sensory data from five operators

There are a total of 150 values that each correspond to an Item (1:10) and Operator (1:5). The first issue is that the variables are stored in both rows (Item) and columns (Operator). The second issue is that the column headers are set as the various Operator treatments, not the variable names (i.e., Operator, Item, and Value). In the tidy dataset, each observation would include an Item and Operator, and a value, giving it a dimension of $(150, 3)$. A third issue is that for i $\in$ {1, 2, $\dots$, 9, 10}, every $(3i-2)$th row has a 6th integer entry on the leftmost side - which should correspond to the Item variable, while the $(3i-1)$th and $3i$th rows only have 5 data entries each. Therefore, the data is manipulated as follows to obtain the targeted dataset with 150 observations of 3 variables.

&nbsp;

```{r, tidy.opts = list(width.cutoff=80), tidy=TRUE}
urla <- "https://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/Sensory.dat"
     # Store URL for sensory data
data_mess <- read.table(urla, skip = 1, header = T, fill = T, sep = " ")
     # Import data; use 2nd row as header
data_tidy <- data_mess
     # Preserve messy data and manipulate using a new variable
for (i in 1:10){
  data_tidy[3 * i - 1, 2:6] <- data_mess[3 * i - 1, 1:5]
     # For each (3*i-1)th row,
     #"push" data to the right to align with rest of the data
     # Do the same for each (3*i)th row
  data_tidy[3 * i, 2:6] <- data_mess[3 * i, 1:5]
}
colnames(data_tidy) <- c("Item", "1", "2", "3", "4", "5")      # Assign Operator numbers
data_tidy <- data_tidy %>% gather(key = Item, value = Value)   # Gather and drop columns
data_tidy <- cbind(data_tidy[, 1], as.data.frame(rep(1:10, each = 3)), data_tidy[, 2])   # Insert Item values into the center column
colnames(data_tidy) <- c("Operator", "Item", "Value")   # Rename variable names
data_tidy$Item <- as.factor(data_tidy$Item)    # Change Item variable from numeric to factor
```
\newpage
```{r}
head(data_tidy, 5)      # Preview sensory data from five operators
summary(data_tidy)   # Summarize sensory data
str(data_tidy)       # Display sensory data structure
```

&nbsp;

b. Gold Medal performance for Olympic Men's Long Jump, year is coded as 1900=0.  

There are a total of 44 values - 22 Year values (-4:92) and 22 long jump records. The first issue is that both the Year and Record variables are listed in multiple columns. The second issue is that the space (" ") separator creates 12 column names for 8 columns of data, although it appears that the column names were intended to be a repeat of "Year" and "Long Jump." Therefore, the data is manipulated as follows to obtain the targeted dataset with 22 observations of 2 variables.

&nbsp;

```{r}
urlb <- "https://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/LongJumpData.dat"
     # Store URL for long jump data
datb_mess <- read.table(urlb, fill = T, header = F, sep = " ", skip = 1)
     # Import data; don't use header because space separator is used
datb_tidy <- datb_mess
     # Preserve messy data and manipulate using a new variable
names(datb_tidy) <- rep(c("Year", "Record"), 4)     # Rename columns
datb_tidy <- as.data.frame(lapply(split(as.list(datb_tidy), names(datb_tidy)), unlist))
     # Split according to names, rejoin into a list of 2, then re-combine into a dataset
datb_tidy <- datb_tidy[ 1:22, 2:1]     # Rearrange columns and remove NA observations
datb_tidy$Year <- as.ordered(datb_tidy$Year)
     # Change Year variable from numeric to ordered factor
head(datb_tidy, 4)      # Preview long jump data
summary(datb_tidy)   # Summarize long jump data
str(datb_tidy)       # Display long jump data structure
```

&nbsp;

c. Brain weight (g) and body weight (kg) for 62 species.

There are a total of 124 values - 62 pairs of brain weights body weights, each corresponding to a mammalian species. The first issue is that both the Brain_g and Body_kg variables are listed in multiple columns. The second issue is that the space (" ") separator creates 12 column names for 6 columns of data, although it appears that the column names were intended to be a repeat of "Brain Wt" and "Body Wt." Therefore, the data is manipulated as follows to obtain the targeted dataset with 62 observations of 2 variables.

&nbsp;

```{r}
urlc <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/BrainandBodyWeight.dat"
     # Store URL for brain and body weight data
datc_mess <- read.table(urlc, fill = T, header = F, sep = " ", skip = 1)
     # Import data; don't use header because space separator is used
datc_tidy <- datc_mess
     # Preserve messy data and manipulate using a new variable
names(datc_tidy) <- rep(c("Brain_g", "Body_kg"), 3)     # Rename columns

datc_tidy <- as.data.frame(lapply(split(as.list(datc_tidy), names(datc_tidy)), unlist))
     # Split according to names, rejoin into a list of 2, then re-combine into a dataset
datc_tidy <- datc_tidy[ 1:62,2:1]     # Rearrange columns and remove NA observations
head(datc_tidy, 5)       # Preview brain weight data
summary(datc_tidy)    # Summarize brain weight data
str(datc_tidy)        # Display brain weight data structure
```

&nbsp;

d. Triplicate measurements of tomato yield for two varieties of tomatos at three planting densities.  

There are a total of 18 tomato yield values, each corresponding to a tomato variety (Ife #1 and Pusa Early Dwarf), and planting density (10,000, 20,000, and 30,000 plants/ha).

&nbsp;

```{r, tidy.opts = list(width.cutoff=80), tidy=TRUE}
urld <- 'http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/tomato.dat'
     # Store URL for brain and body weight data
datd_mess <- fread(urld, header = F, sep = " ")
     # Import data; don't use header
datd_tidy <- datd_mess
     # Preserve messy data and manipulate using a new variable
datd_tidy <- datd_tidy %>% separate(V2, c(paste("10000", 1:3, sep = "_")), sep = ",", remove = T, extra = "drop") %>% separate(V3, c(paste("20000", 1:3,sep="_")), sep = ",", remove = T, extra = "drop") %>% separate(V4, c(paste("30000", 1:3, sep = "_")), sep = ",", remove = T, extra = "drop")
     # Separate data in each column at commas
datd_tidy <- datd_tidy %>% select(-V1)
     # Drop tomato names
datd_tidy <- datd_tidy %>% gather(key = Density, value = Yield)   # Gather and drop columns
datd_tidy <- as.data.frame(cbind(as.character(c(datd_mess[1, 1], datd_mess[2, 1])), as.numeric(rep(c(10000, 20000, 30000), 6)), as.numeric(datd_tidy[, 2])))
     # Re-add Variety and Density variable columns
colnames(datd_tidy) <- c("Variety", "Density", "Yield")    # Rename columns
datd_tidy$Yield <- as.numeric(datd_tidy$Yield)
     # Change Yield variable from ordered to numeric factor
head(datd_tidy, 9)      # Preview tomato data
summary(datd_tidy)   # Summarize tomato data
str(datd_tidy)       # Display tomato data structure
```

## Problem 6

```{r, include = F, tidy.opts = list(width.cutoff=75), tidy=TRUE}
#.datapath <- file.path(path.package('swirl'), 'Courses', 'R_Programming_E',
#                       'Looking_at_Data', 'plant-data.txt')        # Path to data
.datapath <- "/home/ndorphin/R/x86_64-pc-linux-gnu-library/3.4/swirl/Courses/R_Programming/Looking_at_Data/plant-data.txt"
plants <- read.csv(.datapath, strip.white = TRUE, na.strings = "")     # Read in data
.cols2rm <- c("Accepted.Symbol", "Synonym.Symbol")
plants <- plants[, !(names(plants) %in% .cols2rm)]     # Remove annoying columns
names(plants) <- c("Scientific_Name", "Duration", "Active_Growth_Period", "Foliage_Color", "pH_Min", "pH_Max", "Precip_Min", "Precip_Max", "Shade_Tolerance", "Temp_Min_F")
     # Make names pretty
```

```{r}
Foliage_Color_ <- plants$Foliage_Color[!is.na(plants$Foliage_Color) &
      !is.na(plants$pH_Min) & !is.na(plants$pH_Max)]
phmin <- plants$pH_Min[!is.na(plants$Foliage_Color) & !is.na(plants$pH_Min) &
      !is.na(plants$pH_Max)]
phmax <- plants$pH_Max[!is.na(plants$Foliage_Color) & !is.na(plants$pH_Min) &
      !is.na(plants$pH_Max)]     # Exclude rows with missing fcolr, phmin, or phmax values
# Combine 3 columns
plants_tidy <- cbind.data.frame(phmin, phmax, Foliage_Color_)
str(plants_tidy)       # Display tidy plant data structure
# Linear model
model <- lm(formula = phmin + phmax ~ Foliage_Color_, data = plants_tidy)
kable(summary(model)$coef, digits = 3, format = "pandoc",
      caption = "Linear Model Coefficients for Min pH + Max pH ~ Foliage Color")
kable(anova(model), digits = 3, format = "pandoc",
      caption = "ANOVA table for Min pH + Max pH ~ Foliage Color")
```